---
title: "USP 410/510: Urban Informatics"
subtitle: "Visualizing data with ggplot2"
author: "Liming Wang"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    css: "example.css"
    seal: yes
    nature:
      highlightStyle: github
      beforeInit: "macros.js"
---

```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warning=F, echo=F)
options(width = 2000)
options(repos="https://cran.rstudio.com")
```

# ggplot2 - grammar of graphics

`ggplot` is a grammar of graphics, a coherent system for describing and building graphs.

- **data (data frame): ggplot(dataset)**
- **aesthetic mapping (coordination system, color, size, line type, position etc): aes()**
- **geometries: geom_point, geom_line, geom_bar, ...**
- stat (summarization/transformation of data): stat_smooth,
- scale (map data values into computer values): scale_y_continuous(labels=percent), scale_x_log10, ...
- facet: facet_wrap, facet_grid
- fine tuning 
    - coordinate: `coordinate_*`
    - theme: `theme_bw`
    - labels: `labs(x=..., y=...)`

---

# ggplot2 - grammar of graphics

Code template:

```
ggplot(data = <DATA>) +                             # data
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)) +      # aes & geometries
  facet/theme_bw/labs()/...                         # fine tuning
```

---

# Types of Graphics

- numeric variables represent quantities, e.g. time, volume, money, distance, weight
  - histogram
  - scatter plot
- categorical variables represent categories/labels, 
  - nominal categorical: no innate order between values
  - ordinal categorical: with order between values
    - bar plot
- mix & match (choose x & y axis)
  - ridgeline plot (boxplot if you must)
  - scatter plot
  
---

# esquisse

[Interactively explore data by visualizing it with the ggplot2 package](https://github.com/dreamRs/esquisse)

```{r, eval=F, include=TRUE}
install.packages("esquisse")
mtcars <- mtcars %>% 
  mutate(am = as.character(am))
esquisse::esquisser()
```

---

# Interactive Plots
- plotly
- leaflet, tmap for interactive maps

```{r, echo=F}
#install.packages("plotly")
library(ggplot2)
library(plotly)
p <- ggplot(data=mtcars) +
  geom_point(aes(x=wt, y=mpg, size=hp, color=as.factor(gear)))
  
ggplotly(p)
```

---

# ggplot2 - Resources/References

- r graph gallery: https://r-graph-gallery.com/
- ggplot2 cookbook: https://r-graphics.org/
- ggplot2 cheatsheet: https://www.rstudio.com/resources/cheatsheets/
- ggplot2: Elegant Graphics for Data Analysis: https://ggplot2-book.org/
